Improving Haskell Types with SMT

### 2015-06-06 docs/paper.tex
Make changes suggested by reviewers.

@@

SMT solvers, such as CVC4 \cite{cvc4}, Yices \cite{yices}, and Z3
\cite{z3}, implement a wide collection of decision procedures that
work together to solve a common problem.  They have proved to be a useful
tool in both software and hardware verification.
From a user's perspective, the core functionality of an SMT solver is fairly
simple: we may declare uninterpreted constants, assert formulas,
and check if the asserted formulas are {\em satisfiable}.  Checking
for satisfiability simply means that we are asking the
question: are there concrete values for the uninterpreted constants
that make all asserted formulas true.  Here is an example, using
the notation of the SMTLIB standard \cite{smtlib2}

@@

\subsection{Type-Checker Plugins}
In this paper we describe a fairly general extension to the constraint
solver, but other researchers are interested in different extensions,
for example, to add support for units of measure \cite{units-of-measure}.
Instead of having many
ad-hoc extensions directly
in GHC's constraint solver, we collaborated to define and implement an
API for extending GHC's functionality via {\em type-checker plug-ins}.
At present, we are aware of two main
users of this API---our work, and the work on units of measure. We
hope, however, that this mechanism would be useful to other researchers too,
as it makes it fairly easy to experiment with various extensions to
GHC's constraint solver.

@@

The provenance of an equality constraint determines the kinds of
improvements that we can use it for.  Given equalities have solid proof,
and so we may use congruence to rewrite any other constraint.
On the other hand, wanted constraints are goals that need to be proved,
so they {\em cannot} be used to rewrite given constraints.  We may still
use them to rewrite other wanted or derived constraints though.
Finally, derived equalities are implied by the given and wanted constraints
jointly, so they may not be used to rewrite given or wanted constraints
directly, as doing so may lead to circular reasoning.

Instead, derived equalities help with type inference, by guiding the
instantiation of {\em unification variables}. In general,
it is always sound to instantiate a unification variable with whatever
type we want: doing so may reject valid programs, but it will not accept
invalid ones, because we still need to solve all necessary constraints.
Of course, we don't want to reject valid programs, and this is where derived
equalities help us: since they are implied by the goals and assumptions
together, we know that we are not loosing any generality when instantiating
as suggested by a derived constraint.  So, for example, if we compute a
derived constraint \Verb"x ~ Int", and we have a wanted constraint
\Verb"Eq x", and \Verb"x" is a unification variable, then we may
rewrite \Verb"Eq x" to \Verb"Eq Int", which we would proceed to solve
as usual.

@@

Computing new equations is also very important, as it enables collaboration
between the plug-in and the rest of GHC (i.e., the main constraint solver,
and other plug-ins).  For example, consider an implication constraint
of the form:
\begin{Verbatim}
forall x. (x + 5) ~ 8  =>  KnownNat x
\end{Verbatim}
In this case, we'd like to use the SMT solver to compute a new given
equality, \Verb"x ~ 3".  Then, this equality can be used by GHC to
rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3"
which, in turn, can be discharged by the custom solver for the \Verb"KnownNat"
class.  Such collaboration between different solvers is quite common.
Interestingly, this looks a lot like the collaboration between decision
procedures in an SMT solver!  We discuss this observation further
in Section~\ref{modular-typechecker}.

@@

\paragraph{Aside.}
The step of naming foreign types is similar to the flattening step performed
by GHC's constraint solver, the main difference being that GHC names every
call to a type-function, while we name only terms outside of the theory.
Still, this duplication of work is somewhat unsatisfactory, and it would
be nice to extend GHC's flattening pass so that it can be reused by
external plug-ins.

@@

Unfortunately, not all SMT solvers support this feature and CVC4---%
the SMT solver that we used during the development of this work---%
does not have this ability.  To work around this, we implemented
a custom minimization algorithm which work as follows:
\begin{Verbatim}
minimize :: Solver ->
            [Expr] -> -- part of conflict
            [Expr] -> -- search for conflict
            IO [Expr]
minimize s yes [] = return yes
minimize s yes mb =
  do solverPush s
     search s yes [] mb

@@

search s yes mb (c : cs) =
  do solverAssert s c
     res <- solverCheck s
     case res of
       Unsat ->
         do solverPop s
            solverAssert s c
            minimize s (c : yes) mb

@@

The minimization process performs in $O(n^2)$ time, where $n$ is the
number of constraints.  This has not been a big problem,
as this minimization happens only when we report errors and, typically,
the number of constraints is fairly low.

@@

We can compute the values for $A$ and $B$, if we have two examples for
$x$ and two examples for $y$. Note that each call to the solver in
the previous improvement steps gave rise to a potential additional set
of examples, so we have plenty of points to work with: we take care to
remember the values for the variables from two models.   It is also
important that the $x$ values be different, but we can always find such
examples: if we cannot find such an example, then $x$ would have only
one possible value, and it would have been improved already.  It turns
out that this sort of situations occur fairly frequently.  Consider,
for example, the following program fragment:
\begin{Verbatim}
f :: Proxy (a + 1)
f = Proxy

@@

@@

It would be fairly easy to add support for other theories, simply
by declaring additional ``uninterpreted'' kinds and type-functions
in Haskell.  In the rest of this section we explore some of these
possibilities.

@@

While GHC currently does not do any back-tracking search (i.e., decision
procedure may only report inconsistency, or certain knowledge), this
is also the case for Nelson-Open if all theories are convex.  In addition,
there are interesting cases where backtracking might be potentially useful
in Haskell too.  For example, when solving class constraints in the
context of overlapping instances, or instance chains \cite{instance-chains},
it is useful to consider the context of instances,
but back-track and try a different instance, if the context is found
to be unsatisfiable.

>>>>

The approach presented in this paper uses similar tools---namely
SMT solvers---to Liquid Haskell \cite{liquid-haskell}. However,
the design and overall goals of the two techniques are somewhat
different and, indeed, it makes perfect sense to use both systems
in a single project.  Liquid Haskell enables Haskell programmers
to augment Haskell definitions with refinements of the types,
which may be used to verify various properties of the program.
The refinment language is separate from the language of Haskell
types---it is quite expressive, and is really aimed at formal
verification.  As such, a function will have its ordinary
type and, in addition, a refinment type, which serves to specify
some aspect of the behavior of the function.  The checks performed
by Liquid Haskell are entirely separate from the ones performed
by GHC's type checker.  This is quite different from the
technuqe described in this paper, which shows how to integrate
an SMT solver directly into GHC's type chcker.  The benefit of
of extending GHC's type chcker is that we don't need to
introduce a separate language for specifications, instead
we may reuse types directly.  Ultimately, however, the two
approaches are not at all mutually exclusive.  The next
section gives a brief flavor of what can be achieved using
the technieques from this paper.

@@

SMT solvers, such as CVC4 \cite{cvc4}, Yices \cite{yices}, and Z3
\cite{z3}, implement a wide collection of decision procedures that
work together to solve a common problem.  They have proved to be a useful
tool in both software and hardware verification.
From a user's perspective, the core functionality of an SMT solver is fairly
simple: we may declare uninterpreted constants, assert formulas,
and check if the asserted formulas are {\em satisfiable}.  Checking
for satisfiability simply means that we are asking the
question: are there concrete values for the uninterpreted constants
that make all asserted formulas true. The solver may respond in
one of three ways:
\begin{itemize}
\item \Verb"sat", which means that a satisfying assignment exists,
\item \Verb"unsat", which means that a satisfying assignment does not exist,
\item \Verb"unknown", which means that the solver did not find an answer,
      often this is due to reaching some sort of limit (e.g, the solver
      used up too much time).
\end{itemize}

This example uses the notation of the SMTLIB standard \cite{smtlib2}:

@@

\subsection{Type-Checker Plugins}
In this paper we describe a fairly general extension to the constraint
solver, but other researchers are interested in different extensions,
for example, to add support for units of measure \cite{units-of-measure}.

Instead of having many ad-hoc extensions directly in GHC's constraint solver,
we collaborated to define and implement an API for extending GHC's
functionality via {\em type-checker plug-ins}.  At present, we are aware
of three users of this API---our work, a related plug-in that can
solve numerical constraints using custom rewrite rules
\cite{typelits-normalise},
and the work on units of measure \cite{units-of-measure}.

While this infrastructure is still brand new, and not yet stable,
we hope that this mechanism may be useful to other researchers too,
as it makes it fairly easy to experiment with various extensions to
GHC's constraint solver.  Since the framework is aimed at GHC developers,
and is not intended for use in everyday Haskell programming, the plug-ins
have a lot of power.  Indeed, it is not difficult to write a plug-in
that would violate type-safety, in much the same way one could do this
by modifying GHC's constraint solver directly.  The full details of
the type-checker plug-ins API may be found in the GHC User's Guide
\cite{ghc-manual}. Here we discuss only some
of the design choices that we considered, as they have a direct
bearing on the workings of our plug-in.

@@

The provenance of an equality constraint determines the kinds of
improvements that we can use it for.  Given equalities have solid proof,
and so we may use congruence to rewrite any other constraint.
On the other hand, wanted constraints are goals that need to be proved,
so they {\em cannot} be used to rewrite given constraints.  We may still
use them to rewrite other wanted or derived constraints though.
Finally, derived equalities are implied by the given and wanted constraints
jointly, so they may not be used in proofs directly,
as doing so may lead to circular reasoning.

Instead, derived equalities help with type inference, by {\em guiding} the
instantiation of {\em unification variables}. In general,
it is always sound to instantiate a unification variable with whatever
type we want: doing so may reject valid programs, but it will not accept
invalid ones, because we still need to solve all necessary constraints.
Of course, we don't want to reject valid programs, and this is where derived
equalities help us: since they are implied by the goals and assumptions
together, we know that we are not loosing any generality when instantiating
as suggested by a derived constraint.  So, for example, if we compute a
derived constraint \Verb"x ~ Int", and we have a wanted constraint
\Verb"Eq x", and \Verb"x" is a unification variable, then we may
rewrite \Verb"Eq x" to \Verb"Eq Int", which we would proceed to solve
as usual.  This is bit subtle:  a derived equality is never used directly
in a proof---instead it allows us to {\em change what we are trying to prove}
by instantiating unification variables, which may help discharge all
constraints that need solving.

@@

Computing new equations is also very important, as it enables collaboration
between the plug-in and the rest of GHC (i.e., the main constraint solver,
and other plug-ins).  For example, consider an implication constraint
of the form:
\begin{Verbatim}
forall x. (x + 5) ~ 8  =>  KnownNat x
\end{Verbatim}
In this case, we'd like to use the SMT solver to compute a new given
equality, \Verb"x ~ 3".  Then, this equality can be used by GHC to
rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3"
which, in turn, can be discharged by the custom solver for the \Verb"KnownNat"
class\footnote{The class \Verb"KnownNat" is defined in module
\Verb"GHC.TypeLits".  It is used to implement singleton types linking
type-level literals to their run-time representations as \Verb"Integer".}.
Such collaboration between different solvers is quite common.
Interestingly, this looks a lot like the collaboration between decision
procedures in an SMT solver!  We discuss this observation further
in Section~\ref{modular-typechecker}.

@@

\paragraph{Aside.}
The step of naming foreign types is similar to the flattening step performed
by GHC's constraint solver\footnote{GHC's ``flattening'' pass names the
results of all type functions.  This is done before passing constraints
to GHC's internal solver, but the process is reverted before passing the
constraints to the external plug-ins. }.
The main difference being that GHC names every
call to a type-function, while we name only terms outside of the theory.
Still, this duplication of work is somewhat unsatisfactory, and it would
be nice to extend GHC's flattening pass so that it can be reused by
external plug-ins.

@@

Unfortunately, not all SMT solvers support this feature and CVC4---%
the SMT solver that we used during the development of this work---%
does not have this ability.  To work around this, we implemented
a custom reduction algorithm which work as follows:
\begin{Verbatim}
reduce :: Solver ->
            [Expr] -> -- part of conflict
            [Expr] -> -- search for conflict
            IO [Expr]
reduce s yes [] = return yes
reduce s yes mb =
  do solverPush s
     search s yes [] mb

@@

search s yes mb (c : cs) =
  do solverAssert s c
     res <- solverCheck s
     case res of
       Unsat ->
         do solverPop s
            solverAssert s c
            reduce s (c : yes) mb

@@

The reduction process performs in $O(n^2)$ time, where $n$ is the
number of constraints.  This has not been a big problem,
as this reduction happens only when we report errors and, typically,
the number of constraints is fairly low.

Also note that the reduction algorithm may not produce the {\em smallest}
collection of constraints to cause a contradiction.  For example, consider
the constraints $A,B,C$, and suppose that $A,B$ lead to a contradiction
together, and also $C$ causes a contradiction on its own.  Now, depending
on the order in which we process the constraints, we may either compute
$C$ as the result of the reduction, or $A,B$.  This is OK---in this
case the context happens to contain multiple errors, so the reduction
algorithm will simply pick one of them to report.  Of course, a program
would have to eventually fix both errors before the program is accepted.
The important property of the algorithm is that the reduced set does
not contain redundant constraints, in the sense that if we removed
some of the constraints the resulting constraint set would be satisfiable.

@@

We can compute the values for $A$ and $B$, if we have two examples for
$x$ and two examples for $y$.  It is quite standard to ask the SMT
solver for additional examples: this is done by adding additional constraints
that force the models to differ.  In our setting we know that we can
always get another such examples---if this was not the case, a variable
would have exactly one possible values, and we would have improved
it to a constant, as describe previously.  Furthermore, since we do the linear
relation improvement last, we usually can avoid extra calls to the solver
because by this point we typically already have two models where
the variables differ.  It is also
important that the $x$ values be different, but we can always find such
examples: if we cannot find such an example, then $x$ would have only
one possible value, and it would have been improved already.

We admit that this improvement is certainly somewhat bit ad-hoc: after all,
why look for relations between two variables and more?  We implemented
this improvement, because while experimenting with the implementation,
we noticed that this sort of improvement is needed quite often,
at least for a certain class of problems.  So, as far as we know, this
simply a useful heuristic that is worth considering.  It is possible
that alternative techniques (e.g., custom rewrite rules) could achieve
the same effect more directly, and we are certainly considering exploring
such approaches in future work.

To see when this improvement is needed, consider the following program
fragment:
\begin{Verbatim}
f :: Proxy (a + 1)
f = Proxy

@@

\subsection{Our Implementation}
\label{implementation}

To demonstrate the feasibility of the ideas described in this section,
we implemented a GHC plug-in, available from GitHub \cite{type-nats-solver}.
The implementation is intended as a proof-of-concept, and is not
yet ``industrial strength''---more engineering work would be required
to make it more robust and easier to install and configure.

The majority of the effort so far has been focused on understanding how
to integrate an SMT solver with the type-checker, and we have not
conducted extensive testing and measurement of the performance of the plug-in.
It is clear that calling out to an external tool has a cost associated
with it.  However, the preliminary results are encouraging---the plug-in
functions quite well on small examples, but certainly more testing and
experience is needed to find out if the approach will scale to larger
projects.

If despite all the disclaimers, the reader is still interested in
experimenting with the plug-in, we provide the list of GHC options and
extensions that we used in the examples from this paper:

\begin{Verbatim}
{-# OPTIONS_GHC -fplugin=TypeNatSolver #-}
{-# LANGUAGE DataKinds #-}
{-# LANGUAGE TypeOperators #-}
{-# LANGUAGE TypeFamilies #-}
{-# LANGUAGE KindSignatures #-}
{-# LANGUAGE GADTs #-}
\end{Verbatim}

The option \Verb"fplugin=TypeNatSolver" instructs GHC to load the SMT solver
plug-in.  The rest are Haskell extensions that are commonly used in this
setting:
\begin{itemize}
\item \Verb"DataKinds" enables natural number literals at the type
level.
\item \Verb"TypeOperators" and \Verb"TypeFamilies" are need to work
with type-level functions such as \Verb"+".
\item \Verb"KindSignatures" is used to specify the kinds of type variables,
for example, to state that some type-variable is of kind \Verb"Nat".
\item \Verb"GADTs" enables Generalized Algebraic Data-types;  strictly
speaking, these are not required to experiment with the plug-in, however,
many of the more interesting uses of the plug-in end up using GADTs.
\end{itemize}

For the experiments in this paper, we used CVC4, version 1.4.  There was
no deep reason for using this particular solver, other that when we started
working on the project this was the solver with the least restrictive license.
To communicate with the solver, plug-in uses the SMTLIB format (version 2).
Many other SMT solvers support this format (e.g., Z3, Yices), so they
should work just as well.

@@

In principle, it should be fairly easy to add support for other theories,
simply by declaring additional ``uninterpreted'' kinds and type-functions
in Haskell.  Of course---as is often the case---doing so is not completely
trivial, which is why we have not done so yet.  For example, while we
have presented a mechanism for working with the symbols of a new theory
in the type-checker, one still needs to make decisions about which symbols
to import, or what notation to use.  In the rest of this section we
explore some of these possibilities.

@@

While GHC currently does not do any back-tracking search (i.e., decision
procedure may only report inconsistency, or certain knowledge), this
is also the case for Nelson-Open if all theories are convex \cite{NelsonOppen}.
In addition,
there are interesting cases where backtracking might be potentially useful
in Haskell too.  For example, when solving class constraints in the
context of overlapping instances, or instance chains \cite{instance-chains},
it is useful to consider the context of instances,
but back-track and try a different instance, if the context is found
to be unsatisfiable.
